---
title: "Pattern Making Studies"
subtitle: "A Parametric Approach to Flat Pattern Drafting with R"
author: "Your Name"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
---
```{r include=FALSE, cache=FALSE}
# Rscripts/common.R - Atualizado
library(here)

# Criar diretórios necessários
dir.create(here("images", "cache"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output"), showWarnings = FALSE)

# Carregar módulos na ordem correta
source(here("Rscripts", "pattern_core.R"))
source(here("Rscripts", "pattern_notation.R"))
source(here("Rscripts", "pattern_measurements.R"))
source(here("Rscripts", "pattern_blocks.R"))
source(here("Rscripts", "pattern_modifiers.R"))
#source(here("Rscripts", "pattern_area.R"))
source(here("Rscripts", "pattern_plot.R"))
source(here("Rscripts", "pattern_export.R"))
source(here("Rscripts", "pattern_animation.R"))
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
# source("Rscripts/install_packages.R")
```

\frontmatter

# Preface {-}

## About This Book {-}

This book documents my journey learning pattern making through a parametric, code-based approach. Rather than traditional paper drafting, I translate pattern making instructions into R code, creating:

- Reproducible patterns that adjust to any measurement set
- Fully annotated diagrams with all construction points visible
- A searchable reference of drafting techniques
- A record of errors, corrections, and learning insights

## Why Parametric Pattern Making? {-}

Traditional pattern making relies on manual drafting for each size and design. This approach:

1. **Eliminates manual grading**: Change measurements, regenerate the pattern
2. **Preserves design logic**: Every calculation is documented in code
3. **Enables experimentation**: Test design changes instantly
4. **Creates a knowledge base**: Build upon previous work systematically

## Tools Used {-}

- **R** and **R Markdown/Bookdown**: Documentation and computation
- **SeamlyMe**: Body measurement management
- **FreeSewing Notation**: Standard pattern marking symbols
- **Git/GitHub**: Version control and sharing

## How to Use This Book {-}

The book follows a structured learning path:

1. **Chapter 0**: Notation standards and conventions
2. **Chapter 1**: Measurement systems and file formats
3. **Chapters 2a-2f**: Basic blocks for different garments
4. **Chapter 3**: Design variations from basic blocks
5. **Chapter 4**: Real projects with iterations and corrections
6. **Chapter 5**: Reference tables and formulas

## Prerequisites {-}

- R (>= 4.0) and RStudio
- LaTeX distribution (for PDF output)
- Basic understanding of garment construction
- Optional: SeamlyMe for measurement management

## Acknowledgments {-}

- FreeSewing.org for the notation system
- Traditional pattern making textbooks
- Open source community

\mainmatter

<!--chapter:end:index.Rmd-->

```{r include=FALSE, cache=FALSE}
# Rscripts/common.R - Atualizado
library(here)

# Criar diretórios necessários
dir.create(here("images", "cache"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output"), showWarnings = FALSE)

# Carregar módulos na ordem correta
source(here("Rscripts", "pattern_core.R"))
source(here("Rscripts", "pattern_notation.R"))
source(here("Rscripts", "pattern_measurements.R"))
source(here("Rscripts", "pattern_blocks.R"))
source(here("Rscripts", "pattern_modifiers.R"))
#source(here("Rscripts", "pattern_area.R"))
source(here("Rscripts", "pattern_plot.R"))
source(here("Rscripts", "pattern_export.R"))
source(here("Rscripts", "pattern_animation.R"))
```
# Notation Standards and Conventions {#notation-standards}

This chapter defines all notation, terminology, and standards used throughout this pattern making documentation. Following these conventions ensures consistency across all pattern blocks and variations.

## Measurement System

All measurements in this book use the **metric system** exclusively.

| Unit | Abbreviation | Precision | Usage |
|------|-------------|-----------|-------|
| Centimeter | cm | 0.1 | Body measurements, pattern dimensions, ease |
| Millimeter | mm | 0.5 | Notches, seam allowances, small details |
| Meter | m | 0.01 | Fabric length estimation only |

**Critical rule**: Never mix measurement systems. All formulas assume centimeter inputs and return centimeter outputs.

## Coordinate System

All patterns are drafted on a 2D Cartesian coordinate system following textile industry conventions:

- **Origin (0,0)**: Bottom-left corner of the foundation rectangle (point D or C depending on orientation)
- **X-axis**: Horizontal, positive values extend to the right (pattern width)
- **Y-axis**: Vertical, positive values extend upward (pattern length/height)

```
Y-axis (Height)
^
|
|   A(0, height) ----------- B(width, height)
|   |                          |
|   |    FRONT      BACK       |
|   |                          |
|   D(0, 0) --------------- C(width, 0)
|
+------------------------------------------> X-axis (Width)
Origin (0,0)
```

## Point Naming Conventions

Construction points follow traditional pattern drafting conventions with letter designations:

### Primary Structure Points (Foundation Rectangle)
| Point | Description | Typical Position |
|-------|-------------|------------------|
| A | Top-left corner | (0, height) |
| B | Top-right corner | (width, height) |
| C | Bottom-right corner | (width, 0) |
| D | Bottom-left corner | (0, 0) |

### Construction Division Points
| Point | Description | Derivation |
|-------|-------------|------------|
| E, F | Side seam offset | A + offset, D + offset |
| G, H | Center/Armscye division | Midpoint of EB, FC |
| I, J, K, L | Front piece divisions | Divide AG into 4 equal parts |

### Pattern Feature Points
| Category | Points | Represents |
|----------|--------|------------|
| Neckline | N, Decote points | Front and back neckline curves |
| Shoulder | M, P, I, L | Shoulder seam endpoints |
| Armscye | Q, R, S, T, U, V, X, Y, Z | Sleeve cap curve control points |
| Darts | Various | Dart apex and legs |
| Waist/Hip | W, Hip points | Waistline and hip curve points |

## Line Types and Meanings

Based on the [FreeSewing Notation System](https://freesewing.eu/docs/about/notation/), adapted for R/ggplot2:

### Structural Lines
| Line Type | R Linetype | Color | Purpose |
|-----------|-----------|-------|---------|
| **Seam line** | `solid` | Black (#212121) | Actual sewing line |
| **Seam allowance** | `dashed` | Grey (#757575) | Fabric cutting line |
| **Construction line** | `dotted` | Light grey | Temporary guide lines |
| **Grainline** | `dotted` + arrow | Green (#388E3C) | Fabric grain direction |
| **Cut-on-fold** | `twodash` + arrows | Red (#D32F2F) | Place on fabric fold |
| **Center line** | `dashed` | Grey | Center front/back |

### Annotation Lines
| Line Type | R Linetype | Color | Purpose |
|-----------|-----------|-------|---------|
| **Dimension** | `dotted` thin | Blue (#1565C0) | Measurement callouts |
| **Note** | `dashed` thin | Dark grey | Construction notes |
| **Mark** | `solid` thin | Various | Placement indicators |
| **Contrast** | `solid` colored | Cyan (#00BCD4) | Highlight specific areas |
| **Help** | `dotted` thin | Light grey | Assembly guides |

## Notches and Markings

### Notch Types (FreeSewing Standard)
- **Front notch**: Circle with dot center -- indicates front of garment
- **Back notch**: Circle with cross -- indicates back of garment

### Construction Marks
| Symbol | R Implementation | Meaning |
|--------|-----------------|---------|
| (.) | `add_notch(type="front")` | Front orientation |
| (+) | `add_notch(type="back")` | Back orientation |
| (o) | `add_button()` | Button placement |
| (|) | `add_buttonhole()` | Buttonhole placement |
| --x-- | `add_bartack()` | Reinforcement stitch |
| ^ | `add_grainline()` | Grain direction (arrow up) |

## Pattern Piece Information Block

Every pattern piece must include:

1. **Piece identifier**: Number and name (e.g., "1 -- Front Bodice")
2. **Size**: Measurement set identifier
3. **Cutting instructions**: Quantity and fabric type
4. **Date**: Drafting date (ISO format: YYYY-MM-DD)
5. **Scale verification**: 5cm x 5cm test square

## Color Standards

### Fabric Type Colors
| Color Name | Hex Code | R Variable | Usage |
|-----------|----------|------------|-------|
| Primary fabric | #212121 | `fabric_primary` | Main garment fabric |
| Lining | #1976D2 | `fabric_lining` | Lining pieces |
| Interfacing | #F57C00 | `fabric_interfacing` | Interfacing/fusing |

### Annotation Colors
| Purpose | Hex Code | R Variable |
|---------|----------|------------|
| Seam allowance | #757575 | `seam_allowance` |
| Grainline | #388E3C | `grainline` |
| Cut on fold | #D32F2F | `cut_on_fold` |
| Dimensions | #1565C0 | `dimensions` |
| Front notch | #212121 | `notch_default` |
| Back notch | #D32F2F | `notch_back` |

## Grid System

All pattern plots include a reference grid for visual measurement:

- **Major grid lines**: Every 5 cm, solid light grey (#E0E0E0), linewidth 0.3
- **Minor grid lines**: Every 1 cm, very light grey (#F5F5F5), linewidth 0.1
- **Axis labels**: Centered on each major grid line

The grid serves as:
- Quick visual reference for dimensions
- Scale verification tool
- Alignment guide for pattern pieces

## File Naming Conventions

```
{garment}_{block}_{version}_{size}.{ext}

Examples:
  bodice_front_female_v1.0_size48.pdf
  sleeve_basic_v2.1_universal.csv
  skirt_pencil_variation_v1.0_size42_A0.pdf
```

### Version Numbering
- **Major** (X.0.0): Complete redraft or design change
- **Minor** (0.X.0): Significant adjustments to fit
- **Patch** (0.0.X): Small corrections, annotation fixes

## Printing Standards

### Scale Verification
Always print and measure the 5cm test square before cutting. If the square doesn't measure exactly 5cm x 5cm:
1. Check printer settings for "Actual size" (not "Fit to page")
2. Verify PDF scaling is set to 100%
3. Re-print if scale is off by more than 1mm

### Paper Sizes
| Format | Dimensions (cm) | Use Case |
|--------|----------------|----------|
| A4 | 21.0 x 29.7 | Home printing (tiled) |
| A3 | 29.7 x 42.0 | Small patterns |
| A0 | 84.1 x 118.9 | Plotter, full patterns |

### Tile Assembly
For tiled home printing:
1. Trim along registration marks
2. Align matching marks between tiles
3. Tape from center outward
4. Verify scale after assembly

## Measurement File Format

### SeamlyMe CSV Structure
```
code,reference,description,value,formula
height,A01,Height: Total,154,154
bust_circ,G04,Bust circumference,113,113
```

### Ease Parameters CSV Structure
```
parameter,value,unit,description
bust_ease,6,cm,Overall bust ease
shoulder_slope_front,4,cm,Front shoulder drop
```

## R Implementation Notes

### Function Naming
- `calc_*`: Pure mathematical calculations (distance, angle)
- `create_*`: Generate new objects (curves, plots)
- `add_*`: Modify existing plots with annotations
- `read_*` / `write_*`: File I/O operations
- `validate_*`: Check data integrity

### Coordinate Storage
- Points stored as named vectors: `c(x, y)`
- Point collections as named lists: `list(A = c(0, 44), B = c(48, 44))`
- Plotted points as data frames: `data.frame(name, x, y)`

### Cache Strategy
- All plots cached in `images/cache/`
- Cache invalidation based on parameter hash
- Manual cache clear: delete `images/cache/*.pdf`

## References

- FreeSewing Notation Guide: https://freesewing.eu/docs/about/notation/
- ISO 8559: Garment construction and anthropometric surveys
- ABNT NBR 13377: Brazilian standard for garment sizes (where applicable)



<!--chapter:end:00-notation-standards.Rmd-->

```{r include=FALSE, cache=FALSE}
# Rscripts/common.R - Atualizado
library(here)

# Criar diretórios necessários
dir.create(here("images", "cache"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output"), showWarnings = FALSE)

# Carregar módulos na ordem correta
source(here("Rscripts", "pattern_core.R"))
source(here("Rscripts", "pattern_notation.R"))
source(here("Rscripts", "pattern_measurements.R"))
source(here("Rscripts", "pattern_blocks.R"))
source(here("Rscripts", "pattern_modifiers.R"))
#source(here("Rscripts", "pattern_area.R"))
source(here("Rscripts", "pattern_plot.R"))
source(here("Rscripts", "pattern_export.R"))
source(here("Rscripts", "pattern_animation.R"))
```
# Measurement Systems and Integration {#measurements}

This chapter covers how body measurements are captured, stored, and integrated into the parametric pattern drafting system. We use SeamlyMe as the primary measurement tool, but the system accepts any CSV following our format.

## Measurement Philosophy

### Body Measurements vs Pattern Measurements

**Body measurements** are taken directly from the body (or a dress form). They represent the actual dimensions of the person who will wear the garment.

**Pattern measurements** are body measurements plus ease. They represent the dimensions of the finished garment.

Pattern Measurement = Body Measurement + Ease


### Types of Ease

| Ease Type | Purpose | Example |
|-----------|---------|---------|
| **Wearing ease** | Minimum needed for movement and comfort | 4-6 cm at bust |
| **Design ease** | Style choice affecting silhouette | Additional 10 cm for oversized look |
| **Negative ease** | Garment smaller than body (stretch fabrics) | -2 cm at bust for fitted knit |

## SeamlyMe Integration

[SeamlyMe](https://seamly.net/) is open-source software for managing body measurements. It exports measurements in a CSV format we can read directly.

### Export Process from SeamlyMe

1. Open your measurement file (`.vit`) in SeamlyMe
2. Go to **Measurements → Export to CSV**
3. Select all measurements needed for the pattern
4. Save the CSV to `data/measurements/`

### SeamlyMe CSV Format

```csv
code,reference,description,value,formula
height,A01,Height: Total,154,154
bust_circ,G04,Bust circumference,113,113
neck_front_to_waist_f,H01,Neck Front to Waist Front,58,58
```

Columns explained:

    code: Internal identifier used in our R code (e.g., bust_circ)

    reference: SeamlyMe reference code (e.g., G04)

    description: Human-readable description

    value: The actual measurement in centimeters

    formula: Can be a fixed value or a formula referencing other measurements

Required Measurements by Block Type

Female Bodice Block Requirements

# Required measurements for female bodice

```{r, eval=FALSE, echo=TRUE}
female_bodice_required <- c(
  "height",
  "bust_circ",
  "neck_front_to_waist_f",
  "back_waist_length",
  "shoulder_length",
  "armscye_circ",
  "neck_width",
  "across_chest_f",
  "bustpoint_to_bustpoint",
  "bustpoint_to_neck_side"
)
```



# Create reference table
```{r, eval=FALSE, echo=TRUE}
reqs_df <- data.frame(
  Code = female_bodice_required,
  Description = c(
    "Total height",
    "Bust circumference",
    "Front waist length (neck to waist)",
    "Back waist length",
    "Shoulder length",
    "Armscye circumference",
    "Neck width",
    "Across chest (front)",
    "Bust point to bust point",
    "Bust point to neck side"
  ),
  Category = c("Reference", "Circumference", "Length", "Length", 
               "Length", "Circumference", "Width", "Width", 
               "Width", "Length")
)

knitr::kable(reqs_df, 
             caption = "Required measurements for female bodice block",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

Male Bodice Block Requirements

```{r, eval=FALSE, echo=TRUE}
male_bodice_required <- c(
  "height",
  "chest_circ",
  "neck_front_to_waist_f",
  "back_waist_length",
  "shoulder_length",
  "armscye_circ",
  "neck_width",
  "across_back",
  "across_chest_m"
)

reqs_male_df <- data.frame(
  Code = male_bodice_required,
  Description = c(
    "Total height",
    "Chest circumference",
    "Front waist length",
    "Back waist length",
    "Shoulder length",
    "Armscye circumference",
    "Neck width",
    "Across back width",
    "Across chest (male)"
  ),
  Category = c("Reference", "Circumference", "Length", "Length",
               "Length", "Circumference", "Width", "Width", "Width")
)

knitr::kable(reqs_male_df,
             caption = "Required measurements for male bodice block",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Skirt Block Requirements

```{r}
skirt_required <- c(
  "waist_circ",
  "hip_circ",
  "waist_to_hip",
  "waist_to_knee",
  "height"
)

reqs_skirt_df <- data.frame(
  Code = skirt_required,
  Description = c(
    "Waist circumference",
    "Hip circumference",
    "Waist to hip distance",
    "Waist to knee distance",
    "Total height"
  ),
  Category = c("Circumference", "Circumference", "Length", "Length", "Reference")
)

knitr::kable(reqs_skirt_df,
             caption = "Required measurements for skirt block",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Sleeve Block Requirements


```{r}
sleeve_required <- c(
  "armscye_circ",
  "arm_upper_circ",
  "arm_elbow_circ",
  "arm_wrist_circ",
  "arm_shoulder_tip_to_wrist",
  "arm_shoulder_tip_to_elbow",
  "arm_armpit_to_wrist"
)

reqs_sleeve_df <- data.frame(
  Code = sleeve_required,
  Description = c(
    "Armscye circumference",
    "Upper arm circumference",
    "Elbow circumference",
    "Wrist circumference",
    "Shoulder to wrist length",
    "Shoulder to elbow length",
    "Armpit to wrist (inside)"
  ),
  Category = c("Circumference", "Circumference", "Circumference",
               "Circumference", "Length", "Length", "Length")
)

knitr::kable(reqs_sleeve_df,
             caption = "Required measurements for sleeve block",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Pants Block Requirements


```{r}
pants_required <- c(
  "waist_circ",
  "hip_circ",
  "waist_to_hip",
  "crotch_depth",
  "inseam",
  "outseam",
  "thigh_circ",
  "knee_circ",
  "ankle_circ"
)

reqs_pants_df <- data.frame(
  Code = pants_required,
  Description = c(
    "Waist circumference",
    "Hip circumference",
    "Waist to hip distance",
    "Crotch depth (sitting)",
    "Inseam length",
    "Outseam length",
    "Thigh circumference",
    "Knee circumference",
    "Ankle circumference"
  ),
  Category = c("Circumference", "Circumference", "Length",
               "Length", "Length", "Length", "Circumference",
               "Circumference", "Circumference")
)

knitr::kable(reqs_pants_df,
             caption = "Required measurements for pants block",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Loading Measurements in R

The `read_measurements()` Function
```{r}
# Source measurement functions
# source(here::here("Rscripts", "measurement_io.R"))

# Load a measurement file
fem48 <- read_measurements(here::here("data", "measurements", "female_size48.csv"))

# Display key measurements
key_codes <- c("height", "bust_circ", "neck_front_to_waist_f", "shoulder_length")

cat("Key measurements for female size 48:\n")
for (code in key_codes) {
  cat(sprintf("  %-30s: %6.1f cm\n", code, m_get(fem48, code)))
}
```


Creating a Measurement Template

If you don't have SeamlyMe, you can create a template CSV and fill in values manually:

```{r}
# Create template for manual editing
create_measurement_template(
  filepath = "data/measurements/custom_measurements.csv",
  codes = c("height", "bust_circ", "waist_circ", "hip_circ",
            "neck_front_to_waist_f", "back_waist_length",
            "shoulder_length", "armscye_circ", "neck_width")
)
```

Measurement Validation

Checking Required Measurements

```{r}
# Validate that all required measurements are present
tryCatch({
  validate_measurements(fem48, female_bodice_required)
  cat("✓ All required measurements present\n")
}, error = function(e) {
  cat("✗", e$message, "\n")
})

# Check for zero values (likely unfilled)
zero_measurements <- names(fem48)[fem48 == 0]
if (length(zero_measurements) > 0) {
  cat("\n⚠ Zero values found in:", paste(zero_measurements, collapse = ", "))
  cat("\n  These should be measured and filled in before drafting.\n")
}
```

```{r}
# Logical consistency checks
checks <- list()

# Waist should be smaller than bust
if (!is.null(fem48$waist_circ) && !is.null(fem48$bust_circ)) {
  checks$waist_vs_bust <- fem48$waist_circ < fem48$bust_circ
}

# Hip should be larger than waist  
if (!is.null(fem48$hip_circ) && !is.null(fem48$waist_circ)) {
  checks$hip_vs_waist <- fem48$hip_circ > fem48$waist_circ
}

# Arm length checks
if (!is.null(fem48$arm_shoulder_tip_to_wrist) && 
    !is.null(fem48$arm_shoulder_tip_to_elbow)) {
  checks$arm_segments <- fem48$arm_shoulder_tip_to_elbow < 
                          fem48$arm_shoulder_tip_to_wrist
}

cat("Consistency checks:\n")
for (check_name in names(checks)) {
  status <- if (checks[[check_name]]) "✓" else "✗"
  cat(sprintf("  %s %s\n", status, check_name))
}
```




Ease Parameters

Ease parameters are stored separately from body measurements.

This separation allows:

    Changing body measurements without affecting design decisions

    Experimenting with different fits on the same body

    Sharing ease settings across measurement sets

Ease Parameters File Format

```{r, eval=FALSE}
# Read ease parameters
bodice_ease <- read_ease(here::here("data", "parameters", "bodice_ease_female.csv"))

# Display as table
ease_df <- data.frame(
  Parameter = names(bodice_ease),
  Value = paste(unlist(bodice_ease), "cm"),
  Description = c(
    "Overall bust circumference ease",
    "Waist circumference ease",
    "Hip circumference ease",
    "Front shoulder drop from J point",
    "Back neckline drop from B",
    "Reference armscye depth",
    "Front neckline lateral shift",
    "Back neckline lateral shift",
    "Back shoulder dart height",
    "Back shoulder extension",
    "Side seam horizontal offset"
  )
)

knitr::kable(ease_df,
             caption = "Female bodice ease parameters",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```



Ease by Garment Type and Fit
Garment	Close Fit	Semi-Fit	Loose Fit	Oversized
Bodice bust	4-6 cm	6-8 cm	8-12 cm	12-20 cm
Bodice waist	2-4 cm	4-6 cm	6-10 cm	10-16 cm
Skirt hip	2-4 cm	4-6 cm	6-8 cm	8-12 cm
Pants hip	2-4 cm	4-6 cm	6-8 cm	—
Sleeve biceps	2-4 cm	4-6 cm	6-8 cm	8-12 cm
Standard Size Reference Tables
Female Standard Measurements (ABNT NBR 13377)



```{r}
female_standard <- data.frame(
  Size = c("38", "40", "42", "44", "46", "48", "50"),
  Bust = c(84, 88, 92, 96, 102, 108, 114),
  Waist = c(64, 68, 72, 76, 82, 88, 94),
  Hip = c(92, 96, 100, 104, 110, 116, 122),
  Neck_to_Waist_F = c(44, 45, 46, 47, 48, 50, 52),
  Back_Waist = c(38, 39, 40, 41, 42, 43, 44)
)

knitr::kable(female_standard,
             caption = "Standard female measurements (cm) — Reference only",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))

```

Male Standard Measurements

```{r}

male_standard <- data.frame(
  Size = c("46", "48", "50", "52", "54", "56"),
  Chest = c(92, 96, 100, 104, 108, 112),
  Waist = c(78, 82, 86, 90, 96, 102),
  Hip = c(94, 98, 102, 106, 110, 114),
  Back_Waist = c(44, 45, 46, 47, 48, 49)
)

knitr::kable(male_standard,
             caption = "Standard male measurements (cm) — Reference only",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Practical Measurement Tips
Taking Body Measurements

    Use a flexible tape measure — not a metal one

    Measure over undergarments — or the clothing layer the garment will be worn over

    Keep the tape parallel to the floor — especially for circumferences

    Don't pull too tight — the tape should be snug but not compressing

    Mark landmarks — use elastic bands to mark waist, hip, and bust levels

    Take multiple measurements — average three readings for accuracy

    Record posture — note if the subject has forward shoulders, sway back, etc.

Measurement Landmarks
Landmark	How to Find
Waist level	Side bend — where the body creases
Hip level	Widest part of hips, usually 18-22 cm below waist
Bust point	Apex of bust (nipple level)
Neck base	Where neck meets shoulder
Shoulder tip	Acromion bone (bony point at shoulder edge)
Elbow	Olecranon process (bony point of elbow)
Wrist	Styloid process (bony bump at wrist)
Chapter Summary

    All measurements in centimeters, stored in SeamlyMe CSV format

    Measurements are validated for presence and consistency before drafting

    Ease parameters are stored separately for flexibility

    Standard size tables provided for reference only — use actual body measurements when possible

    Each block type has specific measurement requirements documented above





<!--chapter:end:01-measurements.Rmd-->

```{r include=FALSE, cache=FALSE}
# Rscripts/common.R - Atualizado
library(here)

# Criar diretórios necessários
dir.create(here("images", "cache"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output"), showWarnings = FALSE)

# Carregar módulos na ordem correta
source(here("Rscripts", "pattern_core.R"))
source(here("Rscripts", "pattern_notation.R"))
source(here("Rscripts", "pattern_measurements.R"))
source(here("Rscripts", "pattern_blocks.R"))
source(here("Rscripts", "pattern_modifiers.R"))
#source(here("Rscripts", "pattern_area.R"))
source(here("Rscripts", "pattern_plot.R"))
source(here("Rscripts", "pattern_export.R"))
source(here("Rscripts", "pattern_animation.R"))
```
# Curso por Correspondência — Sophia Jobim

**Referência:** JOBIM, Sophia. *O sistema de corte e costura de Sophia Jobim: os anos de ouro de Mme Carvalho no Liceu Império: 1932-1954*. São Paulo: Portal de Livros Abertos da USP, 2024. Disponível em: https://www.livrosabertos.abcd.usp.br/portaldelivrosUSP/catalog/view/1327/1210/4651

Este capítulo documenta o estudo do método de modelagem de Sophia Jobim, transcrevendo as instruções originais e implementando-as em código R parametrizado.

---

## Blusa (Corpo Simples)

**Página de referência:** ~111 do PDF original

### Dados de Entrada

Para um primeiro teste vamos usar as medidas padrão do livro para o manequim 48.

#### Medidas do Manequim 48

```{r carregar-medidas}
# Carregar medidas do arquivo CSV
medidas_sj <- read_measurements(here("data", "measurements", "sophia_jobim_size48.csv"))

# Extrair medidas relevantes
medida_busto <- m_get(medidas_sj, "bust_circ")
comp_frente <- m_get(medidas_sj, "neck_front_to_waist_f")
comp_costas <- m_get(medidas_sj, "back_waist_length")
grossura_braco <- m_get(medidas_sj, "armscye_circ")

# Exibir tabela de medidas
medidas_df <- data.frame(
  Codigo = c("bust_circ", "neck_front_to_waist_f", "back_waist_length", "armscye_circ"),
  Descricao = c("Circunferencia do busto", "Comprimento frente", "Comprimento costas", "Grossura do braco"),
  Valor_cm = c(medida_busto, comp_frente, comp_costas, grossura_braco)
)

knitr::kable(medidas_df, 
             caption = "Medidas do manequim 48 (Sophia Jobim)",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

#### Parametros de Folga e Construcao

```{r carregar-ease}
# Carregar parametros de ease
ease_sj <- read_ease(here("data", "parameters", "sophia_jobim_ease.csv"))

# Exibir tabela de parametros
ease_df <- data.frame(
  Parametro = names(ease_sj),
  Valor = paste(unlist(ease_sj), "cm")
)

knitr::kable(ease_df,
             caption = "Parametros de folga e construcao (Sophia Jobim)",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

#### Tabela de Profundidade da Cava

```{r tabela-cava}
# Carregar tabela de profundidade da cava
tabela_cava <- read_csv(here("data", "measurements", "sophia_jobim_armscye_table.csv"),
                        show_col_types = FALSE)

# Encontrar profundidade para a grossura do braco
profundidade_cava <- tabela_cava$armscye_depth[tabela_cava$arm_circ == grossura_braco]

if (length(profundidade_cava) == 0) {
  profundidade_cava <- approx(tabela_cava$arm_circ, tabela_cava$armscye_depth, 
                              xout = grossura_braco)$y
}

# Destacar a linha usada
tabela_cava$Destaque <- ifelse(tabela_cava$arm_circ == grossura_braco, "<- USADO", "")

knitr::kable(tabela_cava,
             caption = "Tabela 1: Profundidade da cava conforme grossura do braco",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))

cat(sprintf("\n**Braco:** %.0f cm -> **Profundidade da cava:** %.0f cm\n", 
            grossura_braco, profundidade_cava))
```

---

### Instrucoes Originais (transcricao)

> Para facilitar as explicacoes, tomarei como exemplo as medidas do manequim 48, que sao as mais aproximadas do corpo medio feminino. Essas medidas sao, para a blusa: 90 cm para a circunferencia do busto; 44 cm para o comprimento da blusa, na frente, e 41 cm para o comprimento nas costas.
> 
> Para que a blusa nao fique justa como um colete, torna-se necessario dar-lhe um aumento de 6 cm na circunferencia do busto, resultando 96. Como vamos fazer apenas a metade do molde, isto e, metade da frente e metade das costas, dividiremos por 2 a medida total 96 cm. Que resulta 48 cm.

---

### Execucao do Tracado

```{r executar-tracado}
# Executar o tracado com a nova funcao modular
blusa <- draft_blouse_sj(
  medidas = medidas_sj,
  ease = ease_sj,
  armscye_depth = profundidade_cava
)

cat(sprintf("
**Dimensoes do molde:**
- Largura total: %.1f cm
- Altura total: %.1f cm
- Divisao da frente: %.1f cm
- Divisao das costas: %.1f cm
", blusa$dimensions$dim_width, 
   blusa$dimensions$dim_height,
   blusa$dimensions$dim_front_division,
   blusa$dimensions$dim_back_division))
```

---
### Molde Completo

```{r molde-final, fig.cap="Blusa (Corpo Simples) — Sophia Jobim", fig.width=10, fig.height=10}
# Gerar e salvar o gráfico
plot_file <- plot_pattern_notebook(
  block = blusa,
  filename = "blusa_sj_final",
  title = "Blusa (Corpo Simples) — Metodo Sophia Jobim",
  subtitle = paste("Manequim 48 | Busto:", medida_busto, "cm"),
  show_points = TRUE,
  show_grid = TRUE,
  show_construction = TRUE,
  piece_type = "both",
  width = 10,
  height = 10
)

# Incluir a imagem no documento
knitr::include_graphics(plot_file)
```

---

### Comprimento das Linhas

```{r tabela-comprimentos}
comprimentos_df <- data.frame(
  Segmento = c(
    "Ombro frente (I-M)",
    "Decote frente (curva N-I2)",
    "Ombro costas (L-P)",
    "Decote costas (curva B'-L2)",
    "Cava (curva P-M)",
    "Lateral frente (Q-c)",
    "Lateral costas (Q-d)",
    "Cintura frente (c-b)",
    "Cintura costas (d-a)",
    "Barra frente (b-D)",
    "Centro frente (D-N)",
    "Centro costas (a-B')"
  ),
  Comprimento_cm = round(c(
    blusa$line_lengths$shoulder_front_IM,
    blusa$line_lengths$neckline_front,
    blusa$line_lengths$shoulder_back_LP,
    blusa$line_lengths$neckline_back,
    blusa$line_lengths$armscye,
    blusa$line_lengths$side_front_Qc,
    blusa$line_lengths$side_back_Qd,
    blusa$line_lengths$waist_front_cb,
    blusa$line_lengths$waist_back_da,
    blusa$line_lengths$hem_front_bD,
    blusa$line_lengths$center_front_DN,
    blusa$line_lengths$center_back_aB
  ), 1)
)

knitr::kable(comprimentos_df, 
             caption = "Comprimento das linhas que delimitam o molde",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

---

### Area Aproximada para Calculo de Tecido

```{r area-tecido}
area_info <- blusa$area

area_df <- data.frame(
  Peca = c("Frente", "Costas", "Total (2 pecas)"),
  Area_cm2 = c(area_info$front_area_cm2, area_info$back_area_cm2, area_info$total_area_cm2),
  Area_m2 = c(area_info$front_area_m2, area_info$back_area_m2, area_info$total_area_m2)
)

knitr::kable(area_df,
             caption = "Area aproximada das pecas do molde",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"))

cat(sprintf("
**Estimativa de tecido:**
- Area total do molde (frente + costas): %.2f m2
- Com margem de 15%% para curvas e costuras: %.2f m2
- Tecido recomendado (largura 1.40 m): %.2f m linear

> **Nota:** %s
", area_info$total_area_m2,
   area_info$total_area_m2 * 1.15,
   (area_info$total_area_m2 * 1.15) / 1.40,
   area_info$note))
```

---

### Nota Tecnica: Escolha do Metodo para a Curva da Cava

**1. Spline com pontos de controle manuais**

Tentativa de ajustar a curva usando pontos de controle arbitrarios. Resultado: zigue-zague e barriga invertida no decote. Descartado.

**2. Bezier cubica por segmentos**

Dois segmentos (frente: M-Y-T-Q, costas: Q-U-Z-P) com pontos de controle calculados para t=1/3 e t=2/3. Resultado: passou sobre os pontos mas apresentou transicao brusca em Q. Descartado.

**3. LOESS (Local Polynomial Regression)**

Regressao polinomial local com span entre 0.3 e 0.7. Resultado: curvas muito suaves e organicas, mas nao passam exatamente sobre os pontos intermediarios (Z, U, T, Y). Descartado.

**4. Spline cubica parametrica (ESCOLHIDO)**

Spline cubica natural usando distancia acumulada como parametro. Passa exatamente sobre todos os 7 pontos (P, Z, U, Q, T, Y, M). Minimiza a curvatura total, garantindo maxima suavidade matematica. A condicao natural nos extremos permite curvatura zero em P e M. Resolve o problema de ordenacao espacial usando o parametro t baseado em distancia acumulada.

**Decisao:** O metodo escolhido foi a spline cubica parametrica por atender simultaneamente aos dois requisitos: passagem exata sobre todos os pontos de construcao e maxima suavidade da curva.

---

### Exportar para Tamanho Real

```{r exportar-tamanho-real, eval=FALSE}
plot_pattern_fullscale(
  block = blusa,
  output_file = "output/blusa_corpo_simples_A0.pdf",
  paper_size = "A0",
  include_seam_allowance = TRUE,
  seam_allowance = 1.5
)
```

---

### Resumo do Aprendizado

1. **Metodo Sophia Jobim**: usa divisoes proporcionais baseadas no busto
2. **Divisoes**: I e J na frente (1/4 e 3/4 de AG), K e L nas costas (1/4 e 3/4 de GB)
3. **Cava**: construida com spline cubica parametrica sobre 7 pontos (P, Z, U, Q, T, Y, M)
4. **Decote frente**: profundidade = AI + 2 cm, com desvio de 1 cm sobre linha do ombro (ponto I2)
5. **Decote costas**: desce 2 cm de B, com desvio de 1 cm sobre linha do ombro (ponto L2)
6. **Costas**: sempre 3 cm mais estreitas que a frente (deslocamento EF)
7. **Ombro costas**: tem pence de 5 cm em K, prolongada 1 cm para P
8. **Peca das costas**: fecha com pontos a (barra), subindo ate B' (decote costas)
9. **Area aproximada**: calculada por poligono sem curvas; adicionar 15% para margem de costura

<!--chapter:end:02-sophia-jobim.Rmd-->

